---
title: "Final Project Report"
author: "Jiayi Li, Shengkai Zheng, Tianyi Hong"
date: "Dec, 14, 2021"
output: html_notebook
---

## Final Project Report 
* Jiayi Li, Shengkai Zheng, Tianyi Hong
* December, 14, 2021

From 1-20-2021 to 11-10-2021，Which region of the United States (the Northeast, South, West, and Midwest) deals COVID-19 better based on the vaccination rate and Covid-19 cases?


Ever since March 2020, the coronavirus pandemic has affected our lives and nearly every corner of the globe. Coronavirus has also been many people’s concern because of its high transmission rate and fatality rate, compared to other infectious diseases. The infected rate of Coronavirus depends on many elements: contact rate, recovery rate, number of people getting vaccinated, and so on. Therefore, the growth of Coronavirus can vary from country to country, region to region, and even weeks to weeks depending on those elements. In this project, we think it is interesting to investigate the growth of Coronavirus with the number of people getting vaccinated in the U.S, and with numerous data related to coronavirus pandemic, we believe that we can get some meaningful insights. More especially, we came up with the research question: “From 1-20-2021 to 11-10-2021, Which region of United States (the Northeast, South, West, and Midwest) deals COVID-19 better based the vaccination rate and Covid-19 cases?”

The biggest challenge we encountered in this project is the data wrangling part. In this project, we utilize two datasets: StateVaccination, state-by-state data on United States COVID-19 vaccination, and CovidCases, containing aggregate counts of COVID-19 cases and death numbers daily online. To transform these two datasets into the form for visualization and analysis is quite challenging. For example, one of the variables we need is the state that contains states’ names. In the StateVaccination dataset, states’ names are contained in a variable called location. However, some values are invalid for our purpose such as the Bureau of Prisons. To solve this issue, we apply the filter to eliminate all the undesired values. In the CovidCases dataset, the variable state contains all the states’ names. However, states ‘names are in abbreviated forms. To solve that, we use a package called "us data" and leverage the function abbr2state to transform abbreviated names to full names. Another challenge we encountered during the data wrangling part is outliers. There are some outliers in the new_case column in the CovidCases dataset: we notice that for a state like MO and TN, there are some negative values. There is no further information about this attribute provided by the data source that explains about negative values we found. Thus, we took a closer look on our own.

We first visualize the distribution of values in the new_case for each state. Based on the figure “New Covid-19 Cases in States (daily)”, California is among the states which have a lot of negative value for the new case. We try to take a closer look at the negative new_case in California and find an explanation for those negative values. We found that the new_case for 06/30/2021 is negative: -4359. Meanwhile, the total_cases for 06/30/2021 is 3710454, 4359 less than the total_cases for 06/29/2021. Thus, we can conclude that new_case also accounts for people that are recovered from COVID-19. So, we decide to eliminate the negative value and leave the rest like that since the recovery portion is relatively small compared to the infected portion. It will not disturb our results. There are also some minor challenges we encountered in the data wrangling part, but we were able to use different techniques we learned in class to solve those problems. 

Based on the visualizations, we were able to derive some meaningful insights. we found that the overall number of new covid-19 cases in the U.S has been decreased since January, hit the lowest in June, started to increase, and hit the highest point in September. Among the four regions, South seems to have the greatest number of new covid-19 cases in most of the months. There is not a big difference between the other three regions. For the number of people getting vaccinated, people in the South and Mideast are more actively get vaccinated than people in other regions. Overall, the number of people getting vaccinated hit the highest in April and then started to decrease.


### Clean Up Environment and Load Libraries 
```{r}
rm(list = ls())
library(tidyverse)
library(dbplyr)
library(ggplot2)
```


### Data Intake 

```{r}
StateVaccination <- read.csv("state_vaccination.csv")
```

```{r}
CovidCases <- read_csv("United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv")
```

### Inspect data intake 

* Inspect the StateVaccination data frame

The data set StateVaccination is a state-by-state data on United States COVID-19 vaccinations, and is created by *our world in data webiste* which rely on the data updated daily by the United States Centers for Disease Control and Prevention. Their intention is to "maintain the database for the foreseeable future and include additional countries as they implement their vaccination campaigns".This data set is first created on 1/12/2021 and contains data of vaccination related data for each states/federal entity.

```{r}
StateVaccination %>%
  head()
```

```{r}
StateVaccination %>%
    str()
```

```{r}
StateVaccination %>%
    glimpse()
```

* Inspect the CovidCases data frame 

The data set CovidCases is found on DATA.GOV website, and it is created by CDC. CDC reports aggregate counts of COVID-19 cases and death numbers daily online. Data on the COVID-19 website and CDC’s COVID Data Tracker are based on these most recent numbers reported by states, territories, and other jurisdictions. This data set of “United States COVID-19 Cases and Deaths by State over Time” combines this information. However, data are dependent on jurisdictions’ timely and accurate reporting. This data set is first created on November 10, 2020.



```{r}
CovidCases %>% 
  head()

```

```{r}
CovidCases %>%
    str()
```

```{r}
CovidCases %>%
    glimpse()

```




### Data Wrangling and Cleaning 

1. StateVaccination 

* The variable location has some invalid value. Here we eliminate invalid value from location.

```{r}
states <-list("Alabama","Alaska","Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware","Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana","Maine", "Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska", "Nevada", "New Hampshire","New Jersey", "New Mexico", "New York State", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",  "Washington", "West Virginia", "Wisconsin","Wyoming")


StateVaccination <- 
  StateVaccination %>%
  filter( location %in% states)

StateVaccination %>%
  select(location) %>%
  unique() %>%
  head()
```


```{r}
StateVaccination <-
  StateVaccination %>%
  mutate(location = ifelse(location == "New York State", "New York", location))

StateVaccination %>% 
  select(location) %>%
  head()
```

* The variabke date need to be in the interval 2021-1-12 to 2021-11-10

```{r}
library(lubridate)


StateVaccination <-
StateVaccination %>%
  filter(date <= ymd("2021-11-10"))

```


* We create a new variable called month based on date 
```{r}
StateVaccination <-
StateVaccination %>%
  mutate(month = month(date))
  
```


* Created a new variable called peopleVaccinated that contain the total number of vaccinated by the end of each month   

```{r}
CleanedStateVaccination <-
  StateVaccination %>%
  group_by(month, location) %>%
  summarize(peopleVaccinatedMonthly = max(people_vaccinated, na.rm = TRUE)-min(people_vaccinated, na.rm = TRUE))
```


```{r}
CleanedStateVaccination %>% head()
```




2. CovidCases

* We create attribute called month based on submission date 

The variable submission date indicates the date that the data was submitted and recorded of. It is originally character data type, so we first convert it to date date type. Then we create a new attribute called month, taking the month extracted from date.


```{r}
library(lubridate)

```


```{r}
  CovidCases<-
  CovidCases %>%
  mutate(month = str_extract(string = submission_date, pattern = "(^[0-9]{2})")) %>%
  mutate(date = mdy(submission_date)) %>%
  filter(date > mdy("12-31-2020")) %>%
  arrange(submission_date)

CovidCases %>%
    head()
```


```{r}
CovidCases <-
CovidCases %>%
    mutate(month = as.numeric(unlist(month)))

CovidCases %>%
    head()
```

*  transform state attribute to contain full name

In the primary source, location contains full state name. Thus, to facilitate operation on these two data source, we choose to make state attribute to contain full states' name. 

Here we leverage on the abbr2state function  that is provided by the usdata package

```{r}
install.packages("usdata")

```

```{r}
library(usdata)
```

```{r}

CovidCases <-
  CovidCases %>%
  mutate(state = abbr2state(state))

CovidCases %>% 
  select(state) %>%
  unique() %>%
  head()
 
```

From the previous result, we can see that there are rows with null value, and here district of columbia is treated as one state. To stay in accordance with the primary data source, we choose to simply remove rows with state being NA and district of columbia. 

```{r}
states <-list("Alabama","Alaska","Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware","Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana","Maine", "Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska", "Nevada", "New Hampshire","New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",  "Washington", "West Virginia", "Wisconsin","Wyoming")

CovidCases <-
CovidCases %>%
  filter(state %in% states) 

CovidCases %>% select(state) %>%
  unique() %>%
  head()
```

* Here we eliminate negative value in new cases and aggregate it into monthly new cases

```{r}
CovidCases %>%
  ggplot(aes(x= state, y = new_case)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("New Covid-19 Cases in States (daily) ")

```

There are some outliers in the new_case column: we notice that for state like MO and TN, there are some negative value in the new_case column. There is no further information about this attributes provided by the data source. Thus, we take a closer look on our own.

Based on the figure “New Covid-19 Cases in States (daily)”, California is among the states which have a lot of negative value for new case. We try to take a closer look and investigate the negtaive value

from the following result, we can see the new_case for 06/30/2021 is negative: -4359. Meanwhile, the tot_cases for 06/30/2021 is 3710454, 4359 less than tot_cases for 06/29/2021. Thus we can conclude that new_case also account for people that are recovered from COVID-19. So we decide to eliminate the negative value and leave the rest like that since the recover portion is relatively small compared to infected portion. It will not disturb our final results.

```{r}
 CovidCases %>%
   filter(state == "California") %>%
   filter(date == mdy("6-30-2021")| date == mdy("6-29-2021"))
```



```{r}

CovidCases<-
CovidCases %>%
  mutate(new_case = ifelse(new_case < 0, 0, new_case)) 

CovidCases %>%
  filter(new_case < 0)

```

```{r}

CleanCovidCases <-
CovidCases %>%
  group_by(state, month) %>%
  summarize(monthlyNewCases = sum(new_case, na.rm = TRUE))

CleanCovidCases %>%
  head()
```


```{r}
CombinedData <-
CleanedStateVaccination %>%
    rename(state = location) %>%
    inner_join(CleanCovidCases, by = c("month", "state"))
CombinedData %>%
    head()
```


```{r}

northeast<- c("Connecticut","Maine","Massachusetts","New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")

midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota","South Dakota")

south <- c("Delaware", "Florida",  "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "Washington",  "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas")

west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming","Alaska", "California", "Hawaii", "Oregon","Washington")




convertToRegion <- function(data){
  for(i in 1:nrow(data)) {       # for-loop over columns
  if (data[ i,2 ] %in% northeast){
    data[i,2] <-"Northeast"
  }else if (data[ i,2 ] %in% midwest){
    data[i,2] <-"Midwest"
  }else if (data[ i,2 ] %in% south){
    data[i,2] <-"South"
  }else if (data[ i,2 ] %in% west){
    data[i,2] <-"West"
  }
    
    
  }
  return(data)
}

```

```{r}
RegionData = CombinedData


```

```{r}
RegionData = convertToRegion(RegionData)

RegionData

```
```{r}
avgRegionData <-
RegionData %>% 
  group_by(state, month) %>%
  summarize(avgVaccinated = mean(peopleVaccinatedMonthly), avgNewCases = mean(monthlyNewCases))
  
  
avgRegionData

```
### Data Visualization

```{r fig.height=4, fig.width=4}
CleanCovidCases %>%
  ggplot(aes(x = month, y = monthlyNewCases))+
  geom_line(aes(color = state))
  
```


```{r fig.height=4, fig.width=4}

CleanedStateVaccination %>%
  ggplot(aes(x = month, y = peopleVaccinatedMonthly))+
  geom_line(aes(color = location))
  

```

from the above figures, we can see that the number of monthly new cases for most states has dropped from January to June. around the same time (January to mid April, the number of people vaccinated monthly has also been increased. The monthly new COVID-19 cases started to increase from June to mid August, and from September to now (early Nov), the monthly new cases for all states seems to decrease a little. The number of people vaccinated monthly has been decreased for most state from early April to early July, then started to increase around mid-July to August. From August to early-Nov, the number of people getting vaccinated fluctuate a little. 


```{r fig.height=4, fig.width=4}
avgRegionData %>%
  ggplot(aes(x = avgVaccinated, y = avgNewCases))+
  geom_point(aes(shape = state, color = state))+
  facet_wrap(~month)

```


```{r}

avgRegionData %>%
  ggplot(aes(x = month, y = avgNewCases))+
  geom_line(aes(color = state))
  

```

```{r}
avgRegionData %>%
  ggplot(aes(x = month, y = avgVaccinated))+
  geom_line(aes(color = state))

```