---
title: "Final Project Report"
author: "Jiayi Li"
date: "Dec, 14, 2021"
output: html_notebook
---

## Final Project Report 


* Jiayi Li
* December, 14, 2021


#### Narrative Quality (25 pts)

The narrative text (e.g., sentences and paragraphs describing the progression of your analysis) must satisfy the following: 

1. One guiding question or research question for your investigation is clearly stated *with a question mark*
2. Explains why this topic is important and/or interesting to investigate
3. Clearly describes one significant technical challenge encountered during the investigation and how you did or did not overcome it 
4. Explains one or more significant findings or conclusions of your investigation that is **clearly related to the guiding/research question**
5. Is completely free of errors in spelling and grammar



From 1-20-2021 to 11-10-2021，Which region of United States (the Northeast, South, West, and Midwest) deals COVID-19 better based the vaccination rate and Covid-19 cases? 





### Clean Up Environment and Load Libraries 
```{r}
rm(list = ls())
library(tidyverse)
library(dbplyr)
library(ggplot2)
```


### Data Intake 

```{r}
StateVaccination <- read.csv("state_vaccination.csv")
```

```{r}
CovidCases <- read_csv("United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv")
```

### Inspect data intake 

* Inspect the StateVaccination data frame

The data set StateVaccination is a state-by-state data on United States COVID-19 vaccinations, and is created by *our world in data webiste* which rely on the data updated daily by the United States Centers for Disease Control and Prevention. Their intention is to "maintain the database for the foreseeable future and include additional countries as they implement their vaccination campaigns".This data set is first created on 1/12/2021 and contains data of vaccination related data for each states/federal entity.

```{r}
StateVaccination %>%
  head()
```

```{r}
StateVaccination %>%
    str()
```

```{r}
StateVaccination %>%
    glimpse()
```

* Inspect the CovidCases data frame 

The data set CovidCases is found on DATA.GOV website, and it is created by CDC. CDC reports aggregate counts of COVID-19 cases and death numbers daily online. Data on the COVID-19 website and CDC’s COVID Data Tracker are based on these most recent numbers reported by states, territories, and other jurisdictions. This data set of “United States COVID-19 Cases and Deaths by State over Time” combines this information. However, data are dependent on jurisdictions’ timely and accurate reporting. This data set is first created on November 10, 2020.



```{r}
CovidCases %>% 
  head()

```

```{r}
CovidCases %>%
    str()
```

```{r}
CovidCases %>%
    glimpse()

```




### Data Wrangling and Cleaning 

1. StateVaccination 

* The variable location has some invalid value. Here we eliminate invalid value from location by using a filter. 

```{r}

states <-list("Alabama","Alaska","Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware","Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana","Maine", "Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska", "Nevada", "New Hampshire","New Jersey", "New Mexico", "New York State", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",  "Washington", "West Virginia", "Wisconsin","Wyoming")

# eliminate all the invalid value in states
StateVaccination <- 
  StateVaccination %>%
  filter( location %in% states)

# all the unique value in states
StateVaccination %>%
  select(location) %>%
  unique() %>%
  head()
```


```{r}
# here we change New York State to New York.
StateVaccination <-
  StateVaccination %>%
  mutate(location = ifelse(location == "New York State", "New York", location))

StateVaccination %>% 
  select(location) %>%
  head()
```

* The variabke date need to be in the interval 2021-1-12 to 2021-11-10

```{r}
library(lubridate)


StateVaccination <-
StateVaccination %>%
  filter(date <= ymd("2021-11-10"))

```


* We create a new variable called month based on date 
```{r}
StateVaccination <-
StateVaccination %>%
  mutate(month = month(date))
  
```


* We create a Dataframe called CleanedStateVaccination that contain month, location, and a new variable called peopleVaccinated that contain the total number of people get vaccinated each month. Since the variable people_vaccinated is cumulative. We simply group by month and location and subtract the max people_vaccinated and min people_vaccinated to get the number of people get vaccinated each month.


```{r}

CleanedStateVaccination <-
  StateVaccination %>%
  group_by(month, location) %>%
  summarize(peopleVaccinatedMonthly = max(people_vaccinated, na.rm = TRUE)-min(people_vaccinated, na.rm = TRUE))
```


```{r}
CleanedStateVaccination %>% head()
```




2. CovidCases

* We create attribute called month based on submission date 

The variable submission date indicates the date that the data was submitted and recorded of. It is originally character data type, so we first convert it to date date type. Then we create a new attribute called month, taking the month extracted from date.


```{r}
library(lubridate)

```


```{r}
# create variable month 

  CovidCases<-
  CovidCases %>%
  mutate(month = str_extract(string = submission_date, pattern = "(^[0-9]{2})")) %>%
  mutate(date = mdy(submission_date)) %>%
  filter(date > mdy("12-31-2020")) %>%
  arrange(submission_date)

CovidCases %>%
    head()
```


```{r}
# change month to numeric data type 
CovidCases <-
CovidCases %>%
    mutate(month = as.numeric(unlist(month)))

CovidCases %>%
    head()
```

*  transform state attribute to contain full name

In the primary source, location contains full state name. Thus, to facilitate operation on these two data source, we choose to make state attribute to contain full states' name. 

Here we leverage on the abbr2state function  that is provided by the usdata package

```{r}
install.packages("usdata")

```

```{r}
library(usdata)
```

```{r}

CovidCases <-
  CovidCases %>%
  mutate(state = abbr2state(state))

CovidCases %>% 
  select(state) %>%
  unique() %>%
  head()
 
```

From the previous result, we can see that there are rows with null value, and here district of columbia is treated as one state. To stay in accordance with the primary data source, we choose to simply remove rows with state being NA and district of columbia. 

```{r}
states <-list("Alabama","Alaska","Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware","Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana","Maine", "Maryland","Massachusetts","Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska", "Nevada", "New Hampshire","New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia",  "Washington", "West Virginia", "Wisconsin","Wyoming")

CovidCases <-
CovidCases %>%
  filter(state %in% states) 

CovidCases %>% select(state) %>%
  unique() %>%
  head()
```

* Here we eliminate negative value in new cases and aggregate it into monthly new cases

```{r}
CovidCases %>%
  ggplot(aes(x= state, y = new_case)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("New Covid-19 Cases in States (daily) ")

```

There are some outliers in the new_case column: we notice that for state like MO and TN, there are some negative value in the new_case column. There is no further information about this attributes provided by the data source. Thus, we take a closer look on our own.

Based on the figure “New Covid-19 Cases in States (daily)”, California is among the states which have a lot of negative value for new case. We try to take a closer look and investigate the negtaive value

from the following result, we can see the new_case for 06/30/2021 is negative: -4359. Meanwhile, the tot_cases for 06/30/2021 is 3710454, 4359 less than tot_cases for 06/29/2021. Thus we can conclude that new_case also account for people that are recovered from COVID-19. So we decide to eliminate the negative value and leave the rest like that since the recover portion is relatively small compared to infected portion. It will not disturb our final results.

```{r}
 CovidCases %>%
   filter(state == "California") %>%
   filter(date == mdy("6-30-2021")| date == mdy("6-29-2021"))
```



```{r}

CovidCases<-
CovidCases %>%
  mutate(new_case = ifelse(new_case < 0, 0, new_case)) 

CovidCases %>%
  filter(new_case < 0)

```

* Here we create CleanCovidCases DataFrame that contain state, month, and monthlyNewCases. 
```{r}

CleanCovidCases <-
CovidCases %>%
  group_by(state, month) %>%
  summarize(monthlyNewCases = sum(new_case, na.rm = TRUE))

CleanCovidCases %>%
  head()
```

* Here we create dataframe called CombinedData by joining CleanedStateVaccination and CleanCovidCaes.
```{r}
CombinedData <-
CleanedStateVaccination %>%
    rename(state = location) %>%
    inner_join(CleanCovidCases, by = c("month", "state"))
CombinedData %>%
    head()
```

* Here we classify each state into four region: northeast, midwest, south, and west using the self-defined function: convertToRegion
```{r}

northeast<- c("Connecticut","Maine","Massachusetts","New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania")

midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota","South Dakota")

south <- c("Delaware", "Florida",  "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "Washington",  "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas")

west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming","Alaska", "California", "Hawaii", "Oregon","Washington")




convertToRegion <- function(data){
  for(i in 1:nrow(data)) {       # for-loop over columns
  if (data[ i,2 ] %in% northeast){
    data[i,2] <-"Northeast"
  }else if (data[ i,2 ] %in% midwest){
    data[i,2] <-"Midwest"
  }else if (data[ i,2 ] %in% south){
    data[i,2] <-"South"
  }else if (data[ i,2 ] %in% west){
    data[i,2] <-"West"
  }
    
    
  }
  return(data)
}

```

```{r}
RegionData = CombinedData


```

```{r}
RegionData = convertToRegion(RegionData)

RegionData

```
* Then we get the avgVaccinated and avgNewCases for each region in each month.
```{r}
avgRegionData <-
RegionData %>% 
  group_by(state, month) %>%
  summarize(avgVaccinated = mean(peopleVaccinatedMonthly), avgNewCases = mean(monthlyNewCases))
  
  
avgRegionData

```
### Data Visualization

1. Monthly New Covid-19 cases for each state through Jan to Nov

```{r fig.height=4, fig.width=4}
CleanCovidCases %>%
  ggplot(aes(x = month, y = monthlyNewCases))+
  geom_line(aes(color = state))+
  ggtitle("Monthly New Covid-19 cases for each state through Jan to Nov")
  
```

2. Number of people get vaccinated each month for each state through Jan to Nov

```{r fig.height=4, fig.width=4}

CleanedStateVaccination %>%
  ggplot(aes(x = month, y = peopleVaccinatedMonthly))+
  geom_line(aes(color = location)) +
  ggtitle("Number of people get vaccinated each month for each state through Jan to Nov")
  

```

3. The average number of new covid-19 cases and average number of people get vaccinated for each US region from Jan to Nov.

```{r fig.height=4, fig.width=5.5}
avgRegionData %>%
  ggplot(aes(x = avgVaccinated, y = avgNewCases))+
  geom_point(aes(shape = state, color = state))+
  facet_wrap(~month)+
  ggtitle("average number of new covid-19 cases and average number of people get vaccinated for each US region from Jan to Nov")

```

4. average number of new covid-19 cases for each US region from Jan to Nov
```{r}

avgRegionData %>%
  ggplot(aes(x = month, y = avgNewCases))+
  geom_line(aes(color = state))+
  ggtitle("average number of new covid-19 cases for each US region from Jan to Nov")
  

```
5. The average number of people get vaccinated for each US region from Jan to Nov.


```{r}
avgRegionData %>%
  ggplot(aes(x = month, y = avgVaccinated))+
  geom_line(aes(color = state)) +
  ggtitle("The avg number of people get vaccinated for each US region from Jan to Nov.")

```
6. Here we use US map to better visualize the monthly new covid-19 cases for each state on June (lowest monthly new covid-19) and September(highest monthly new covid-19) for each state. 


```{r}
library(mosaic)

CleanCovidCases%>%
  filter(month == 6) %>%
  mUSMap(key = "state", fill = "monthlyNewCases") 


```

```{r}

CleanCovidCases%>%
  filter(month == 9) %>%
  mUSMap(key = "state", fill = "monthlyNewCases")  

```